Delocalization of Eigenvectors in Random Matrices

### 2025-03-23 Draft.tex
stochastic domination

\begin{equation*}
\begin{array}{c}
G_1 = G\left(H_{11}, \cdots, H_{ij}^G, \cdots, H_{NN}^G\right)\\
G_2 = G\left(H_{11}, \cdots,\tilde{t}_{ij}^G H_{ij}^G=\cdots H_{NN}^G\right) \\
G_3 = G\left(H_{11}, \cdots,\tilde{t}_{ij} H_{ij}, \cdots, H_{NN}^G\right),
\end{array},
\end{equation*}
where $\tilde{t}_{ij}^G$ and $\tilde{t}_{ij}$ are the replacement variables at step $ij$ for this argument, different than the ones we used in our Lindeberg argument. By the matrix identity $(2.4)$ for $A = H_1-z$ and $B = H_2-H_1$, we have the following: 
$$G_2 = G_1 - G_1\left(H_2-H_1\right)G_2$$
$$\Rightarrow G_1 = G_2+G_1\left(H_2-H_1\right)G_2$$

\noindent Given that $H_2$ and $H_1$ differ only in their $(ij)$th and $(ji)$th component, we can represent their difference as the sum of two matrices with the only non-zero entries being the latter, which by definition of the orthonormal basis is $e_i^*e_j$ and $e_j^*e_i$ respectively. Then, for $c_{12} = \tilde{t}_{ij}^G-1$, we have:
\begin{equation*}
G_1 = G_2 + G_1\left((cH_{ij}^Ge_i^*e_j+ cH_{ji}^G e_j^*e_i\right)G_2
\end{equation*}

$$\Rightarrow (G_2)_{xy} = (G_1)_{xy} + cH_{ij}^G (G_1)_{xi}(G_2)_{jy}+\tilde{c}H_{ij}^G (G_1)_{xj}(G_2)_{iy}$$
Given $(8.1.2)$ and $(8.1.3)$, we have:
$$|(G_2)_{xy}|\prec 1 + W^{-\frac{1}{2}}|(G_2)_{jy}|+W^{-1/2}|(G_2)_{iy}|$$
Let $\Xi = \max_{x, y}|(G_2)_{xy}|$, then: $$\Xi \prec 1 + W^{-1/2}\Xi,$$
hence $\Xi\prec 1$ (\textcolor{red}{requires justification for the $N^\epsilon$ exponents from the stochastic domination}. 
\textcolor{red}{complete the continuity argument}.

@@

$$\lesssim W^{-3+\epsilon} + W^C W^{-3} W^{-D_0}$$
We can pick $D_0$, s.t. the stochastically dominating term is $W^{-3+\epsilon}$, s.t. we get $\mathcal{P}+\mathcal{P}^G\lesssim W^{-3+\epsilon}$, i.e: $$\left\vert \mathbb{E}\left(F(H_{ij}\right) - \mathbb{E}\left(F(H_{ij}^G\right)\right\vert\lesssim W^{-3+\epsilon},$$
Then, by applying the triangle inequality to the telescoping sum of replacements, and accounting for the $NW$ non-zero entries in the band, we have: 
 \begin{equation*} \left\vert \mathbb{E}\left(F(H)\right) - \mathbb{E}\left(F(H^G)\right)\right\vert\lesssim NWW^{-3+\epsilon} = NW^{-2+\epsilon}\tag{8.35?}
 \end{equation*}
\noindent For the condition of delocalization, we need for any $\beta>0$ the following to hold:
 $$W\gg N^{\frac{1}{2}+\beta} = N^{\frac{1+2\beta}{2}}\Leftrightarrow N\ll W^{\frac{2}{1+2\beta}}$$
\noindent Bounding $(8.3)$ by the latter condition gives us:
  $$NW^{-2+\epsilon}\ll W^{-2+\epsilon+\frac{2}{1+2\beta}}$$
\noindent Hence, in order for our asymptotics to work, we need a negative exponent, i.e: $$-2+\epsilon + \frac{2}{1+2\beta}<0$$
  $$0<\epsilon <2-\frac{2}{1+2\beta} = \frac{4\beta}{1+2\beta}>0,$$
  whereas since $\beta>0\Rightarrow \frac{4\beta}{1+2\beta}$, we can always pick such an $\epsilon>0$ by the density of the rationals. This is also why we needed a five matching moments assumption - if we had four instead, then the bound we would get per condition $(8.1.2)$ would be $\left\vert \mathbb{E}\left(F(H_{ij}\right) - \mathbb{E}\left(F(H_{ij}^G\right)\right\vert\lesssim W^{-\frac{5}{2}+\epsilon}$. By following the same steps as above, we would get $-\frac{3}{2}+\epsilon+\frac{2}{1+2\beta}<0\Leftrightarrow \epsilon< \frac{6\beta-1}{1+2\beta}$.  But then $\epsilon$ would be negative for $\beta<\frac{1}{6}$, meaning the delocalization bound does not hold.

>>>>

\begin{equation*}
\begin{array}{c}
G_1 = G\left(H_{11}, \cdots, H_{ij}^G, \cdots, H_{NN}^G\right)\\
G_2 = G\left(H_{11}, \cdots,\tilde{t}_{ij}^G H_{ij}^G=\cdots H_{NN}^G\right) \\
G_3 = G\left(H_{11}, \cdots,\tilde{t}_{ij} H_{ij}, \cdots, H_{NN}^G\right),
\end{array},
\end{equation*}
where $\tilde{t}_{ij}^G$ and $\tilde{t}_{ij}$ are the replacement variables $\in [0, 1]$ at step $ij$ for this argument, different than the ones we used in our Lindeberg argument. By the matrix identity $(2.4)$ for $A = H_2-z$ and $B = H_1-H_2$, we have the following: 
$$G_1 = G_2 - G_2\left(H_1-H_2\right)G_1$$
$$\Rightarrow G_2 = G_1+G_2\left(H_1-H_2\right)G_1$$

\noindent Given that $H_2$ and $H_1$ differ only in their $(ij)$th and $(ji)$th component, we can represent their difference as the sum of two matrices with the only non-zero entries being the latter, which by definition of the orthonormal basis is $e_i^*e_j$ and $e_j^*e_i$ respectively. Then, for $c = 1-\tilde{t}_{ij}^G\in [0, 1]$, we have:
\begin{equation*}
G_2= G_1 + G_2\left(cH_{ij}^Ge_i^*e_j+ cH_{ji}^G e_j^*e_i\right)G_1
\end{equation*}

$$\Rightarrow (G_2)_{xy} = (G_1)_{xy} + cH_{ij}^G (G_2)_{xi}(G_1)_{jy}+{c}H_{ij}^G (G_2)_{xj}(G_1)_{iy}$$
Given $(8.1.2)$ and $(8.1.3)$, we have:
$$|(G_2)_{xy}|\prec 1 + W^{-\frac{1}{2}}|(G_2)_{jy}|+W^{-1/2}|(G_2)_{iy}|$$
Let $\Xi_2 = \max_{x, y}|(G_2)_{xy}|$. We can rewrite the expression above as: 
\begin{equation*}\Xi \prec 1 + W^{-1/2}\Xi_2\tag{8.7}\end{equation*}

\noindent Now, let $\epsilon, D>0$ be arbitrary. Given that we work in the context of the delocalization, we have $W\gg N^{\frac{1}{2}+\beta} = N^{\frac{1+2\beta}{2}}\Rightarrow W^{-1/2}\ll N^{-\frac{1+2\beta}{4}}$. Hence, let us choose $\delta = \min \left(1, \frac{2\beta+1}{4\epsilon}\right)$. 
By Def.1 of stochastic domination, from $(8.7)$, we have that: $$\sup_{s\in S_N}\mathbb{P}\left(\Xi_N(s)>N^{\delta\epsilon}(1+W^{-1/2}\Xi_N(s))\right)<N^{-D}$$
$$\sup_{s\in S_N}\mathbb{P}\left(\Xi_N(s)>N^{\delta\epsilon}+N^{\delta\epsilon} W^{-1/2}\Xi_N(s)\right)<N^{-D}$$
$$\sup_{s\in S_N}\mathbb{P}\left(\Xi_N(s)(1-N^{\delta\epsilon} W^{-1/2})>N^{\delta\epsilon}\right)<N^{-D}$$
$$\sup_{s\in S_N}\mathbb{P}\left(\Xi_N(s)>\frac{N^{\delta\epsilon}}{(1-N^{\delta\epsilon} W^{-1/2})}\right)<N^{-D}$$
By our choice of $\delta$, we clearly have that $$N^{\delta\varepsilon}W^{-1/2} \ll N^{\delta\varepsilon} \cdot N^{-1/4-\beta/2} = N^{\delta\varepsilon-1/4-\beta/2} \ll 1$$
This means that $\exists N_0(\beta, \epsilon)$, s.t $\forall N\geq N_0$, we have that $\frac{N^{\delta\epsilon}}{(1-N^{\delta\epsilon} W^{-1/2})}>N^{\delta\epsilon}$. This means that for all $N\geq N_0$, the latter event is a subset: $$\left\{\Xi_N(s)>\frac{N^{\delta\epsilon}}{(1-N^{\delta\epsilon} W^{-1/2})}\right\}\subset \left\{\Xi_N(s)>N^{\delta \epsilon}\right\}$$
Define

hence $\Xi_2\prec 1$.
Notice that the approach is equivalent for $\Xi_3 = \max_{x, y}|(G_3)_{xy}|$, i.e WLOG, we have that $\Xi_3\prec 1$ as well.

\textcolor{red}{complete the continuity argument}.

@@

$$\lesssim W^{-3+\epsilon} + W^C W^{-3} W^{-D_0}$$
We can pick $D_0$, s.t. the stochastically dominating term is $W^{-3+\epsilon}$, s.t. we get $\mathcal{P}+\mathcal{P}^G\lesssim W^{-3+\epsilon}$, i.e: $$\left\vert \mathbb{E}\left(F(H_{ij}\right) - \mathbb{E}\left(F(H_{ij}^G\right)\right\vert\lesssim W^{-3+\epsilon},$$
Then, by applying the triangle inequality to the telescoping sum of replacements, and accounting for the $NW$ non-zero entries in the band, we have: 
 \begin{equation*} \left\vert \mathbb{E}\left(F(H)\right) - \mathbb{E}\left(F(H^G)\right)\right\vert\lesssim NWW^{-3+\epsilon} = NW^{-2+\epsilon}\tag{8.8}
 \end{equation*}
\noindent For the condition of delocalization, we need for any $\beta>0$ the following to hold:
 $$W\gg N^{\frac{1}{2}+\beta} = N^{\frac{1+2\beta}{2}}\Leftrightarrow N\ll W^{\frac{2}{1+2\beta}}$$
\noindent Bounding $(8.8)$ by the latter condition gives us:
  $$NW^{-2+\epsilon}\ll W^{-2+\epsilon+\frac{2}{1+2\beta}}$$
\noindent Hence, in order for our asymptotics to work, we need a negative exponent, i.e: $$-2+\epsilon + \frac{2}{1+2\beta}<0$$
  $$0<\epsilon <2-\frac{2}{1+2\beta} = \frac{4\beta}{1+2\beta}>0,$$
  whereas since $\beta>0\Rightarrow \frac{4\beta}{1+2\beta}$, we can always pick such an $\epsilon>0$ by the density of the rationals. This is also why we needed a five matching moments assumption - if we had four instead, then the bound we would get per condition $(8.1.2)$ would be $\left\vert \mathbb{E}\left(F(H_{ij}\right) - \mathbb{E}\left(F(H_{ij}^G\right)\right\vert\lesssim W^{-\frac{5}{2}+\epsilon}$. By following the same steps as above, we would get $-\frac{3}{2}+\epsilon+\frac{2}{1+2\beta}<0\Leftrightarrow \epsilon< \frac{6\beta-1}{1+2\beta}$.  But then $\epsilon$ would be negative for $\beta<\frac{1}{6}$, meaning the delocalization bound does not hold.
